{\LARGE \textsc{Biotic invasions reduce the manageability of mutualistic networks}}

### 2016-05-12 paper/manuscript.Rnw
edit introduction

Alternative titles:

@@

\textit{Tentative spiel for the first paragraph:} We need to manage ecological communities because of ecosystem services and biodiversity maintenance. We would like to control communities in the sense of conserving and restoring them. However to control them (in that sense) we need to understand how species affect each other and potentially the whole system when they are faced by perturbations or management interventions. But we cannot do that yet because we lack a theoretical framework that link the dynamics of the community to management. Control theory might provide this link.

In contrast with other types of complex systems like neural, intra-organizational, or trust neworks, ecological networks do not have a hierarchical or distributed structure but rather tend to be closed systems in which resources recirculate\todo{add ref}. As such, ecosystems contain multiple feedbacks that allow them to self-regulate and, therefore, function relatively independent from external stimulation\todo{add ref}.

Research stemming from control theory has established a strong link between the structure of complex networks and their controlability and that, in principle it is possible to alter a ecological community's composition by modifying the abundances of only a few species \autocite{Liu2011, Cornelius2013}. Based on this work and using a small number of binary food-webs (only presence or absence of interactions), it has been suggested that the degree distribution of ecological networks makes them inherently difficult to control \autocite{Liu2011, Ruths2014}. This theoretical result is in agreement with many cases in which the management of ecological communities is more challenging than aticipated.  Although these conceptual advances on control theory have provided an initial link between the structure of ecological communities and our ability to manage them, this relationship is still ambiguous. Ultimately, the challenge of ecological networks is represented by the variablity of their structure and interaction strengths---which can vary by several orders of mangitude\todo{add ref}.

For instance, biotic invasions, as other major drivers of global change, can induce dramatic changes on the patterns of interactions that determine the structure of ecological networks \autocite{Baxter2004, Tylianakis2008, Ehrenfeld2010a}. These changes can be particularly pronounced in mutualistic networks of plants and pollinators where biotic invasions have been shown to modify the strength of species interactions and the degree of network nestedness and connectivity \autocite{Olesen2002, Aizen2008, Bartomeus2008, Vila2009, Traveset2013}. Moreover, as evidenced by the limited amount of success of restoration projects \autocite{Smith2016}, returning invaded communities to a predisturbance state is quite a difficult endeavour \autocite{Suding2004, Rodewald2015}. Understanding how invasion-induced differences in network structure impact our ability to manage them, can provide useful lessons to conservation science. What is more, there are also ecological insights to be had: by explicitly taking into account the extent to which changes in the abbundances of one species may ripple trough the community, applying control theory to ecological networks can provide an indication of which species are important from a dynamic perspective, and therefore crucial when attempting to alter (or maintain) the ecosystem state.

\textcite{Isbell2013} used a control theoretic approach to find the minimum subset of species neccessary to maintain network structure in six marine food webs. They found that humans affect a larger proportion of this subset of species than, for instance, apex predators. They argue that this disproportionate influence could translate into unintentional restructuring of food webs. Although insightful, the scope and applicability of \textcite{Isbell2013} can be extended in numerous ways. First, although they find the size of this minimum set, they disregard the fact that the set species composition is not unique and thus not all species that might be included are equally important for the community population dynamics. Second, their methods were naive to differences on the strength of interspecific effects. And third, their approach was conceptually limited to trophic interactions\todo{is this too harsh?}.

Here, we expand previous theory of the control of complex systems and use a set of paired invaded and uninvaded plant-pollinator communities to investigate the link between network structure and our ability to manage them. Plant-pollinator networks provide an ideal framework to answer these questions. On one hand, community networks that quantify relative levels of interaction are readily available. On the other, the bipartite nature of pollination networks makes it is possible to simplify assumptions of how these interactions translate into interspecific effects. Our theoretical extension of network control, allow us to ask the question of whether biotic invasions as drivers of change in network structure increasses or decreasses the manageability of communities as well as the relative importance of species at driving the population dynamics of other species in the community. \todo{punchy line to finish here?}.

@@

>>>>

Alternative title:

@@

Complex systems, such as ecological communities, are characterised by non-linear relationships in which the whole is greater than the sum of its parts \todo{add refs}. This complex systems approach has led to the development of a rich suite of analytical and simulation tools with which to understand the role of species when embedded in an intricate network of interactions\todo{add refs}. However, in contrast to other types of complex systems like neural, intra-organizational, or trust neworks, ecological networks do not have a hierarchical or distributed structure but rather tend to be closed systems in which resources recirculate\todo{add ref}. As such, ecosystems contain multiple feedbacks that allow them to self-regulate and, therefore, function relatively independent from external stimulation\todo{add ref}. This complex propertes have hindered, or at least complicated, our ability to find management solutions to the many problems ecological communities face. A framework that allow us to explain, predict, and manage ecological communities, particularly when they are faced with perturbations, needs to be able to account for their complex structure, and the dynamics that determine the state of each species---and the community as a whole. Control theory might be a good candidate framework as it is designed to deal with determining and supervising the behavior of a dynamical system and is well equiped to deal with the many feedbacks presents in ecological communities\todo{add ref}.

Research stemming from control theory has established a strong link between the structure of complex networks and their controlability, and that, in principle, it is possible to alter a ecological community's composition by modifying the abundances of only a few species \autocite{Liu2011, Cornelius2013}. Based on this work, it has been suggested that the degree distribution of ecological networks makes them inherently difficult to control \autocite{Liu2011, Ruths2014}. This theoretical result is in agreement with many examples in which the management and restoration of ecological communities is more challenging than aticipated. For instance, biotic invasions, as a major drivers of global change, can induce dramatic changes on the patterns of interactions that determine the structure of ecological networks \autocite{Baxter2004, Tylianakis2008, Ehrenfeld2010a}. The difficulties of returning invaded communities to a predisturbance state can be evidenced by the limited amount of success of restoration projects \autocite{Suding2004, Rodewald2015, Smith2016}. Although it is a double challenge, understanding how differences in network structure before and after the invasion impact our ability to manage them is a first step towards an informed recovery.

A control theoretical approach not only might provide insight to conservation but also for ecological science: by explicitly taking into account the extent to which changes in the abbundances of one species may ripple trough the community, applying control theory to ecological networks can provide an indication of which species are important from a dynamic perspective, and therefore crucial when attempting to alter (or maintain) the ecosystem state. Nevertheless, although it has been prevoiously used to highlight the disproportionate influence that humans have on marine food-webs \autocite{Isbell2013}, control theory has has rarely been aplied in an ecological context. One reason might be that the control of complex networks is still a young field and the tools necessary to account for variation on the strength of interspecific effects, which are characteristic of ecological networks, are not there.

Here, we expand previous theory of the control of complex systems and use a set of ten paired invaded and uninvaded plant-pollinator communities to investigate the link between network structure and our ability to manage them. We focus on biotic invasions because they can produce tractable changes in the structure of ecological networks. These changes can be particularly pronounced in mutualistic networks of plants and pollinators where biotic invasions have been shown to modify the strength of species interactions and the degree of network nestedness and connectivity \autocite{Olesen2002, Aizen2008, Bartomeus2008, Vila2009, Traveset2013}. In addition plant-pollinator networks provide an ideal framework to answer these questions. On one hand, community networks that quantify relative levels of interaction are readily available. On the other, the bipartite nature of pollination networks makes it is possible to simplify assumptions of how these interactions translate into interspecific effects. Specifically we ask two key questions framed in the context of a plant invasion. First, we quantify the manageability of pollination communities. Second, we ask whether some species are more important than others at driving the population dynamics of the community and which factors determine this importance.

\iffalse

@@

\fi
